---
title: "Evaluating undercounts in epidemics"
output:
  pdf_document
date: "`r format(Sys.Date(), '%d %B %Y')`"
author: "Michael Li, Jonathan Dushoff, David J. D. Earn, and Ben Bolker"
bibliography: undercount.bib
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{grffile}
urlcolor: blue
---

## Introduction

Two papers (@bohningEstimating2020 and @maruottiEstimating2022) have promoted a formula that claims  to estimate the completeness of sampling of an infectious disease epidemic, using only case reporting counts, based on a mark-recapture formula developed by Chao. We believe this formula is fundamentally wrong, and that it is impossible to estimate undercounting without some kind of auxiliary information.

From @bohningEstimating2020:

> The idea is to apply [the Chao estimator] day-wise. We take an arbitrary day $t$. At this day we have $\Delta N(t)$ new infections. This will be viewed as $f_1$, the infected people identified just once. If we look at $\Delta N(t - 1)$, then this is the count of new infections the day before. But these will still be infected at day $t$ unless they decease. So, $f_2$ [the number of infections counted twice] corresponds to $\Delta N(t - 1) - \Delta D(t)$. We can ignore the number of recoveries as we are looking at infections which are very recent (notified at day $t$ or $t - 1$). Hence we are able to give the estimate for the number of hidden infections at day $t$ as $H(t) = \frac{[\Delta N(t)]^2}{\Delta N(t - 1) - \Delta D(t)}$ \ldots

Or from @maruottiEstimating2022:

> We will denote with $N(t)$ the cumulative count of infections at week $t$ where $t = t_0, \dots, t_m$. Hence $\Delta N (t) = N(t) -N(t-1)$ are the number of  new  infections at week $t$ where $t = t_0, t_1,  \dots t_m$ ... Here $\Delta N(t)$ corresponds to the infected people identified just once, and $\Delta N(t-1)$  is the  number  of  those  identified  twice. 

This latter paper does not use a death term, and gives the full bias-corrected expression (their eq 1) as

\[
\frac{\Delta N(t) (\Delta N(t) - 1)}{1 + \Delta N(t-1)} ;
\]

the bias-correction ($\pm 1$) terms don't affect the critique below.

## Critique

This approach is not an appropriate application of these mark-recapture formulas. Why are cases identified at time $t-1$ representative of the number of cases counted twice? The fact that the same individual *could* be counted twice in the cumulative case report (for some sampling designs) is not relevant. How can comparing yesterday's count to today's provide information about the completeness of sampling, or anything other than the rate of growth of the epidemic? 

In principle, it is possible to estimate the number of unobserved individuals (hidden cases) if individuals can be re-identified when observed a second time, or even with unmarked individuals given an appropriate sampling design [@royleHierarchical2008]. 
<!--- However, public health case reporting is typically unrelated to mark-recapture (although it can be used to estimate completeness of contact tracing: @lerdsuwansriSensitivity2022, @polonskyNovel2022). Furthermore, --> In practice public health case reporting does not use such sampling designs. Case reporting is usually exclusive (i.e. someone who has been identified as a case will not be reported again later) and/or anonymized so that we cannot identify whether a particular case/infected individual was double-counted or not.

### Exponential example

Suppose an epidemic is growing exponentially. In the epidemic phase, the various measures of current epidemic size (incidence, prevalence, cumulative incidence, etc.) all grow exponentially, or geometrically if we sample in discrete time. Suppose $I(t)$ ("new cases" or incidence) is growing at a geometric rate $\lambda$ per time step, i.e. $I(t) = I(0) \lambda^t$, and suppose we are sampling a fraction $a$ (for "ascertainment") of the incidence. Again for simplicity, we'll compute the simple/non-bias-corrected estimate, without death or recovery:

$$
\begin{split}
H(t) & = \frac{[\Delta N(t)]^2}{\Delta N(t - 1)} \\
& = \frac{a^2 I(0)^2 \lambda^{2t}}{a I(0) \lambda^{t-1}} \\
& = a I(0) \lambda^{t+1} \\
& = a \lambda I(t).
\end{split}
$$

This shows that the estimated number of hidden counts depends on the growth rate $\lambda$ as well as the ascertainment fraction $a$ --- not what we wanted. The "undercount ratio" $H(t)/\Delta N(t) = H(t)/(a I(t)) = \lambda$ is independent of the ascertainment fraction.

Furthermore, the estimated total incidence is:
$$
\begin{split}
\textrm{inc}(t) & = \Delta N(t) + H(t) \\
& = aI(t) + a \lambda I(t) \\
& = aI(t)(1+\lambda) 
\end{split}
$$
so the estimated ratio of incidence to observed cases $\textrm{inc}(t)/\Delta N(t) = (1+ \lambda)$ is again independent of $a$. 

More generally, looking at the form of the expression $[\Delta N(t)]^2/\Delta N(t - 1)$, we can see that the undercount will always be equal to the rate of acceleration of observed cases ($\Delta N(t)/\Delta N(t-1)$) times the number of new cases $\Delta N(t)$, so undercounting will always be assumed to be large (proportional to observed cases) when disease incidence is increasing, and small when incidence is decreasing.

### Simulation example (SIR model)

```{r get_pars, echo=FALSE}
library(shellpipes)
loadEnvironments()
options(digits = 2)
pct <- function(x) round(100*x)
R0  <- round(beta/gamma,2)
inf_per <- round(1/gamma,2)
```

We simulate a simple SIR epidemic in discrete time, apply two different ascertainment fractions to derive a time series of case reports, and apply the authors' formula (including bias corrections) to estimate the number of hidden cases.
At the peak of the epidemic, when the number of cases is large, the method predicts the same ratio of hidden cases to observed case reports regardless of the ascertainment fraction; at the beginning and end of the epidemic, the bias correction makes underreporting appear lower in the lower-ascertainment scenario, when the number of reported cases is lower.

<!-- note N is hard-coded, too hard to format nicely/automatically -->
```{r get_plot, echo = FALSE, out.height = "4in", fig.cap = sprintf('Simulated SIR epidemic with ${\\cal R}_0 = %1.1f$, infectious period of %1.2f days, $N = 10^6$. Left panel, cumulative numbers of reported and estimated-hidden cases along with true incidence (only one line appears because both scenarios have the same true incidence). Right panel, ratio of estimated (black) or true (blue) hidden cases to reported cases for each scenario.', R0, inf_per)}
knitr::include_graphics(matchFile(exts="pdf"))
```

<!-- We simulate a simple SIR epidemic with a constant ascertainment rate `r 100*c_prop`% (i.e., an  underreporting level of $(1-a)/a \cdot 100 = `r pct((1-c_prop)/c_prop)`$%), and apply the authors' formula. 
-->

We conclude that the authors' suggested formula should not be applied to disease outbreak incidence data.

\newpage

## References

